---
title: "BO-NLME_Tutorial"
author: "Jim Hughes"
date: "2026-01-18"
output: 
  html_document: 
    toc: true
editor_options: 
  markdown: 
    wrap: 72
---

# Set up

```{r, echo = TRUE, warning = FALSE, message = FALSE}
# Clear working directory
  rm(list = ls(all = TRUE))

# Load package libraries
  library(conflicted)
  library(tidyverse)
  library(grid)
  library(gridExtra)
  library(GGally)
  library(scales)

# Handle package conflicts
  conflict_scout()
  conflicts_prefer(dplyr::filter, dplyr::lag, purrr::discard, 
    readr::col_factor, gridExtra::combine)
  
# Source custom functions
  source("var_labels.R")
  source("model_summary.R")
  source("plot_diagnostics.R")
  source("plot_vpc.R")

# Set directory paths
  dat_dir <- "../data"
  scr_dir <- "../scripts"
  mod_dir <- "../models"
  out_dir <- "../output"
  
# Set ggplot2 theme
  theme_bw2 <- theme_set(theme_bw(base_size = 12))
  theme_update(plot.margin = unit(c(1, 1.5, 1, 1), "lines"))
```

# Data

The dataset used for this vignette comes from the clinical trial
described below. Only the data from the control arm is included in the
dataset.

-   <https://clinicaltrials.gov/study/NCT00081796?tab=table>
-   <https://www.sanofi.com/assets/dotcom/content-app/clinical-studies/pharma/Letter-L/EFC6089_summary.pdf>

```{r, echo = TRUE, warning = FALSE, message = FALSE}
# Load raw data
  raw_path <- file.path(dat_dir, "EQLQ_Clean_2004_23JUN2025.csv")
  raw_data <- read_csv(raw_path, col_types = cols(
    .default = col_double(),
    UID = col_character(),
    NCT = col_character(),
    QSFLAG = col_character(),
    SEX = col_character(),
    MENOS = col_character(),
    ECOG = col_character(),
    CHEMO = col_character(),
    SURGERY = col_character(),
    RADIO = col_character(),
    ERS = col_character(),
    PGRS = col_character(),
    MHDEPRESSION = col_character(),
    MDANXIETY = col_character()
  ))
  
# Define covariate types
  yesno_covs <- c("CHEMO", "SURGERY", "RADIO", "MHDEP", "MDANX")
  posneg_covs <- c("ERS", "PGRS")
  cat_covs <- c("STDID", "SEX", "RACE", "MENOS", "ECOG", "DTH", yesno_covs, posneg_covs)
  cont_covs <- c("AGE", "LESION1", "HEIGHT", "WEIGHT", "DTHDY", "PFSDY")
```

The raw data has a mix of integers and character strings describing
different aspects of the study. This is cleaned up prior to preparing
the data for anlaysis.

```{r, echo = TRUE, warning = FALSE, message = FALSE}
# Prepare analysis dataset
  clean_data <- raw_data %>%
    rename(TAFD = TIME, MHDEP = MHDEPRESSION, MDANX = MDANXIETY) %>%
  # Map items to their respective scales
    left_join(unnest(qs_scales, FLAG), by = "FLAG") %>%
  # Convert categorical covariates to number format
    mutate(SEX = as.integer(factor(SEX, levels = c("Male", "Female")))) %>%
    mutate(MENOS = as.integer(factor(MENOS, 
      levels = c("premenopausal", "other")))) %>%
    mutate(ECOG = as.integer(factor(ECOG, levels = c("0", "1", "3-Feb"))) - 1L) %>%
    mutate(across(all_of(yesno_covs), \(x) if_else(x == "Y", 1, 0))) %>%
    mutate(across(all_of(posneg_covs), \(x) if_else(x == "Positive", 1, 0))) %>%
    mutate(DTH = na_if(DTH, -999))
```

For bounded outcomes analysis you will be working with aggregate scores,
but how you put those scores together depends on what approach you want
to use. For this example the QLQ-C30 scores are calculated for
visualization and presentation, along with a simple sum of scores which
will be used for analysis as it can be used for both continuous and
discrete bounded outcomes methods.

```{r, echo = TRUE, warning = FALSE, message = FALSE}
  score_data <- clean_data %>%
  # Set flag to determine which scores require imputation
    group_by(across(all_of(c("ID", "TAFD", "TAFR", "SCALE", 
      cat_covs, cont_covs)))) %>%
    mutate(IMPFL = length(DV) != n_items) %>%
  # Calculate scores
    summarise(
      raw_score = mean(DV),  # missing items imputed if at least 1 item remains
      score = if_else(unique(is_func),
        true = 100*(1 - (raw_score - 1)/unique(range)),  # Functioning
        false = 100*(raw_score - 1)/unique(range)  # Global Health + Symptom
      ),
      sum_score = sum(DV),
      raw_score_noimp = if_else(unique(IMPFL), NA_real_, raw_score),
      score_noimp = if_else(unique(IMPFL), NA_real_, score),
      sum_score_noimp = if_else(unique(IMPFL), NA_real_, sum_score),
      NQS = length(unique(FLAG)),
      .groups = "drop"
    ) %>%
  # Convert to long format and convert to numeric flag
    pivot_longer(contains("score"), names_to = "FLAG", values_to = "DV") %>%
    mutate(FLAG = as.integer(factor(FLAG, 
      levels = c("raw_score", "score", "raw_score_noimp", 
        "score_noimp", "sum_score", "sum_score_noimp")
    ))) %>%
    mutate(C = NA_real_, .before = 1)
```

## Explore data

Exploration of data prior to analysis is just as important for bounded
outcomes analysis as other NLME anlayses. Typical visualizations used
for exploring categorical or continuous data can be used.

For this vignette, the focus will be on the Physical Functioning domain
of the questionnaire, due to it having a sufficient range of values to
justify the use of beta regression (n = 16).

A key exploratory plot to explore is the distribution of the dependent
variable. This can help determine whether the chosen method needs to
handle values on the boundary or not. Below simple summarisation and
histograms are used to explore the distribution, and identify that the
data has a reverse J-shaped distribution, with many values at the lower
boundary and some values at the upper boundary.

```{r, echo = TRUE, warning = FALSE, message = FALSE}
  plot_data <- score_data %>%
    filter(FLAG == 1 & SCALE == 2) %>%
    left_join(select(qs_scales, SCALE, label), by = "SCALE") %>%
    mutate(label = factor(label, levels = qs_scales$label)) %>%
    reduce2(.init = ., catcov_labels, cat_covs, \(init, cat, cov) {
      left_join(init, cat, by = cov)
    }) %>%
    mutate(CFB = DV - first(DV), .by = "ID")
  count(plot_data, lower_bound = DV == 1, upper_bound = DV == 4)
  p <- ggplot(data = plot_data)
  p <- p + geom_histogram(aes(x = DV), bins = 16, 
    fill = "#2BA7FF", colour = "black")
  p

```

Simple individual plots are shown as another example, to help visualize
the types of PRO trajectories the final model will need to describe. The
data has varied profiles, with an overall indication that a reasonable
starting point for model development is a structural model that can
describe some kind of placebo/standard of care effect along with disease
progression.

```{r, echo = TRUE, warning = FALSE, message = FALSE}
  id_data <- plot_data %>%
    mutate(UID = paste0("STDID: ", STDID, ", ID: ", ID))
  unique_id <- unique(id_data$UID)
  unique_flag <- unique(id_data$label)
  id_plots <- map(unique_id, \(id) {
    plot_data <- filter(id_data, UID == id & FLAG == 1)
    pal_index <- which(unique_flag %in% unique(plot_data$label))
    p <- ggplot(data = plot_data)
    p <- p + geom_point(aes(x = TAFD, y = DV, colour = label))
    p <- p + geom_line(aes(x = TAFD, y = DV, colour = label))
    p <- p + labs(x = "Dependent Variable", y = "Time After First Dose (days)")
    p <- p + facet_wrap(c("UID", "label"), scales = "free", ncol = 3)
    p <- p + scale_shape_manual("Prediction", values = c(1, 2))
    p <- p + scale_linetype_manual("Prediction", values = c("solid","dashed"))
    p <- p + theme(legend.position = "none", 
      plot.margin = unit(c(0.3, 0.3, 0.3, 0.3), "cm"),
      axis.title.x = element_text(size = 10), 
      axis.text.x = element_text(size = 10),
      axis.title.y = element_text(size = 10), 
      axis.text.y = element_text(size = 10)
    )
  })
  plot_grid <- marrangeGrob(grobs = id_plots, ncol = 3, nrow = 2,
    npages = ceiling(length(id_plots)/6))
  grid.draw(plot_grid[[2]])
```

# Model Definition

## Zero-/One-inflated Beta Regression

Prior to defining the model, the dependent variable $DV$ needs to be
transformed such that it is constrained between 0 and 1. As PROs tend to
already bounded (lowest score 1, highest score 15), the boundaries
simply need to be shifted.

$$ y=\frac{DV-DV_{lo}}{DV_{up}-DV_{lo}} $$

where $DV$ represents the raw integer score, while $DV_{lo}$ and
$DV_{hi}$ represent the lower and upper boundary of the raw sum of
scores respectively.

The main difference between beta regression and traditional PK/PD
modelling is the assumption of the dependent variable being produced
from a beta distribution rather than a normal distribution. Therefore
the regression only differs in the residuals, with the probability that
the value $y$ is observed being described by:

$$ p(y)=Beta(\alpha,\beta) \\ p(y)=\frac{\Gamma(\alpha+\beta)}{\Gamma(\alpha)\Gamma(\beta)}y^{\alpha-1}(1-y)^{\beta-1} $$

where $\alpha$ and $\beta$ are shape parameters for the beta
distribution and $y$ is constrained between 0 and 1.

A more convenient parameterization of the beta regression is to consider
it in terms of the mean and variance of the distribution, such that:

$$ E[Y]=\mu \\ Var[Y]=\frac{\mu(1-\mu)}{\phi+1} \\ \alpha=\mu\phi \\ \beta=(1-\mu)\phi $$

where $\mu$ is the mean prediction (constrained between 0 and 1) and
$\phi$ is free to control the residuals (similar to $\sigma$ for a
normal distribution.

To handle observations at the boundary, an inflated beta distribution is
necessary such that:

$$ p(y) = \begin{cases} p_0 & \text{if }y=0 \\ p_1 & \text{if }y=1 \\ Beta(\alpha,\beta)(1-p_0-p_1) & \text{otherwise} \end{cases} \\ p_0=\text{logit}^{-1}(-\zeta_1-\zeta_2\cdot\text{logit}(\mu))\\ p_1=\text{logit}^{-1}(-\zeta_1+\zeta_2\cdot\text{logit}(\mu)) $$

where the inflation terms $p_0$ and $p_1$ are defined by a pair of
functions with some decent properties, but most importantly that force
the condition $p_0+p_1\le1$.

The remainder of the model remains largely the same as a typical PK/PD
or exposure response model, ensuring that the resulting prediction can
be transformed to produce a value constrained between 0 and 1.

$$\text{logit}(\mu)=BASE+P_{eff}+D_{prog}$$

where $IPRED$ is equal to baseline, $BASE$, plus placebo effect,
$P_{eff}$, and disease progression, $D_{prog}$, and can take values from
$-\infty$ to $\infty$.

## Discretized Beta Regression

Rather than using the beta probabiliity density function to determine
the probability of observation $y$, instead discretized beta regression
uses the cumulative distribution function. It does this by splitting the
beta distribution up into $n$ segments, determining the probability of
being between $y_{lo}$ and $y_{up}$.

$$ y_{lo}=\frac{DV-DV_{lo}}{DV_{up}-DV_{lo}+1}\\ y_{up}=\frac{DV-DV_{lo}+1}{DV_{up}-DV_{lo}+1}\\ p(y)=P(y_{lo}<Y\le y_{hi})=I_{y_{hi}}(\alpha,\beta)-I_{y_{lo}}(\alpha,\beta)  $$

where $y_{lo}$ and $y_{hi}$ represent the breakpoints between all
possible values of $y$, and $I_y$ is the cumulative distribution
function for a beta distribution.

Many aspects of beta regression apply to discretized beta regression,
including the ability to parameterize as either $\alpha$ & $\beta$ or
$\mu$ & $\phi$, and ability to use standard structural models. The main
difference and a potential advantage of discretized beta regression is
that presence of observations on the boundary do not necessitate
zero-/one-inflation, as the lowest and highest segments represent the
boundaries.

# Model Runs and Diagnostics

For this vignette, two initial models are presented as a starting point.
As with all non-linear mixed effects analysis, model development is an
iterative process guided by an overall goal. These models are reasonable
for producing EBEs representing PRO trajectories for use as predictors
for other outcomes, but further work would be necessary to produce
models with the goal of prediction.

Two models are provided in this vignette, the first "run7" analyzes the
PRO data as a continuous bounded outcome using beta regression. The
second analysis "run9" analyzes the PRO data as a discrete bounded
outcome using discretized-beta regression. It should be noted that the
objective function values between these two methods are not directly
comparable, with inflated beta regression treating the data as
continuous and using densities to calculate the likelihood, while
discretized beta regression treats the data as discrete and uses
probabilities to calculate the likelihood.

## Zero-/One-inflated Beta Regression

On looking at the model code in `models/run7/run7.mod` you may notice
two things. First, the in-built beta function is used for efficiency.
These functions take a vector `VBT` that has the observation in the
first position, and $\alpha$ and $\beta$ in the second position.
Secondly, there is no disease progression component in this version of
the model, as there was difficulty in minimization when including this
parameter in a addition to the two parameters handling zero- and
one-inflation.

```{r, echo = TRUE, warning = FALSE, message = FALSE}
  run7_path <- file.path(mod_dir, "run7")
  run7_summary <- model_summary(path = run7_path, param_labels = TRUE)
  run7_summary$ObjectiveFunctionValue
  run7_summary$ConditionNumber
  run7_summary$Theta
  run7_summary$Omega
```

For diagnostics individual predictions, population predictions and
residuals are needed. These are all relatively simple to calculate for
an inflated beta regression and are included as part of the control
stream.

This vignette looks at goodness of fit and individual plots, but all
standard NLME diagnostics can be applied here.

```{r, echo = TRUE, warning = FALSE, message = FALSE}
# Read in subject-level data
  run7_eta <- file.path(run7_path, "run7.eta") %>%
    read_table(na = c("NA", "***********", "1.#INFE+00"), skip = 1, 
      col_types = cols(.default = col_double()))
# Read in observation-level data
  run7_fit <- file.path(run7_path, "run7.fit") %>%
    read_table(na = c("NA", "***********", "1.#INFE+00"), skip = 1, 
      col_types = cols(.default = col_double())) %>%
    left_join(run7_eta, by = "ID") %>%  # Add subject level data
    mutate(PROJ = "PF2") %>%
  # Add in categorical covariate information
    reduce2(.init = ., catcov_labels, cat_covs, \(init, cat, cov) {
      left_join(init, cat, by = cov)
    })
```

```{r, echo = TRUE, warning = FALSE, message = FALSE, fig.dim = c(8, 8)}
# Create goodness of fit plots
  ipred_dv_plot <- obs_gof_plot(run7_fit, 
    pred_name = "IPRED", 
    dv_name = "DV", 
    cov_name = "PROJ",
    pred_label = "Individual Predicted")
  pred_dv_plot <- obs_gof_plot(run7_fit, 
    pred_name = "PPRED", 
    dv_name = "DV",
    cov_name = "PROJ",
    pred_label = "Individual Predicted")
  wres_pred_plot <- res_gof_plot(run7_fit, 
    res_name = "PWRES", 
    x_name = "PPRED",
    cov_name = "PROJ", 
    x_label = "Population Predicted",
    res_label = "Weighted Residual")
  wres_time_plot <- res_gof_plot(run7_fit, 
    res_name = "PWRES", 
    x_name = "TAFD",
    cov_name = "PROJ", 
    units = "hours", 
    x_label = "Time After First Dose",
    res_label = "Weighted Residual")
# Render plot
  combine_gof_plot(ipred_dv_plot, pred_dv_plot, 
    wres_pred_plot, wres_time_plot)
```

```{r, echo = TRUE, warning = FALSE, message = FALSE}
# Create individual plots
  run7_id_data <- mutate(run7_fit, SID = paste0("ID: ", ID))
  p <- individual_plots(run7_id_data, 
    dv_name = "DV",
    ipred_name = "IPRED",
    pred_name = "PPRED",
    time_name = "TAFD",
    id_name = "SID",
    flag_name = "PROJ",
    plots_per_page = 6
  )
  grid.draw(p[[2]])
```

Simulations can be performed in R for a simple model such as this. However,
this vignette will be using NONMEM, with simulation code provided in 
`models/run7/sim7.mod`. `.fit` and `.eta` files have been pre-processed and 
saved to binary file to reduce size of vignette material.

```{r, echo = TRUE, warning = FALSE, message = FALSE}
# Prepare data for visual predictive check
  obs_data <- score_data %>% 
    filter(SCALE == 2 & FLAG == 5) %>%
    mutate(PROJ = "PF2") %>%
    mutate(DV = 100*(1 - (DV/NQS - 1)/3))  # convert to score
  # run7_simeta <- file.path(run7_path, "sim7.eta") %>%
  #   read_table(na = c("NA", "***********", "1.#INFE+00"), skip = 1,
  #     col_types = cols(.default = col_double()))
  # run7_sim <- file.path(run7_path, "sim7.fit") %>%
  #   read_table(na = c("NA", "***********", "1.#INFE+00"), skip = 1,
  #     col_types = cols(.default = col_double())) %>%
  #   left_join(run7_eta, by = "ID") %>%  # Add subject level data
  #   filter(!is.na(SIM)) %>%
  #   mutate(PROJ = "PF2") %>%
  #   reduce2(.init = ., catcov_labels, cat_covs, \(init, cat, cov) {
  #     left_join(init, cat, by = cov)
  #   })
  # saveRDS(run7_sim, file.path(mod_dir, "run7", "sim7.rds"))
  run7_sim <- readRDS(file.path(mod_dir, "run7", "sim7.rds")) %>%
    mutate(PROJ = "PF2") %>%
    mutate(DV = 100*(1 - (DV/NQS - 1)/3))  # convert to score
  vpc_data <- list(obs = obs_data, sim = run7_sim) %>%
    map(mutate, WAFD = TAFD / 7)
  
# Define VPC bins
  tafd_labels <- c(0, 4, 8, 12, 16, 24, 40, 56)*7
  tafd_breaks <- c(-Inf, head(tafd_labels, -1) + diff(tafd_labels) / 2, Inf)
# Bin data for VPC
  tadbin_data <- vpc_data %>%
    map(bin_times, 
      time_var = "WAFD", 
      breaks = tafd_breaks, 
      labels = tafd_labels
    )
  
# Summarise and plot VPC for percent change from baseline
  vpc_summary <- tadbin_data %>%
    map(summarise_vpc,
      time_var = "TIMEBIN",
      facet_cov = "PROJ")
  plot_vpc(
    obs_summary = vpc_summary$obs,
    sim_summary = vpc_summary$sim,
    obs_data = tadbin_data$obs,
    time_var = "WAFD",
    time_label = "Time After First Dose (weeks)",
    dv_label = "Physical Functioning Score\n",
    png_name = "VPC_run7"
  )
  
```

## Discretized Beta Regression

Similar to the inflated beta regression example, the model code in
`models/run9/run9.mod` uses the in-built beta CDF function for
efficiency. For this model the disease progression component was able to
be included with successful minimization. It is possible this is due to the more
parsimonious form of discretized beta regression.

```{r, echo = TRUE, warning = FALSE, message = FALSE}
  run9_path <- file.path(mod_dir, "run9")
  run9_summary <- model_summary(path = run9_path, param_labels = TRUE)
  run9_summary$ObjectiveFunctionValue
  run9_summary$ConditionNumber
  run9_summary$Theta
  run9_summary$Omega
```

Again, individual predictions, population predictions and residuals are
needed. This is a little more difficult with discretized beta
regression, so in this example they are calculated outside of NONMEM.

```{r, echo = TRUE, warning = FALSE, message = FALSE}
# Read in subject-level data
  run9_eta <- file.path(run9_path, "run9.eta") %>%
    read_table(na = c("NA", "***********", "1.#INFE+00"), skip = 1, 
      col_types = cols(.default = col_double()))
# Read in observation-level data
  run9_fit <- file.path(run9_path, "run9.fit") %>%
    read_table(na = c("NA", "***********", "1.#INFE+00"), skip = 1, 
      col_types = cols(.default = col_double())) %>%
    left_join(run9_eta, by = "ID") %>%  # Add subject level data
    mutate(PROJ = "PF2") %>%
  # Add in categorical covariate information
    reduce2(.init = ., catcov_labels, cat_covs, \(init, cat, cov) {
      left_join(init, cat, by = cov)
    }) %>%
  # Calculate predictions (i.e. expectation) and residuals
  # Requires determining the probability of each possible value y can take
  # Population predictions
    mutate(PRED = pmap_dbl(list(PMU, DVLO, DVUP, PHI), 
      .f = \(PMU, DVLO, DVUP, PHI) {
        reduce(.init = 0, seq(DVLO, DVUP), \(init, x) {
          xlo <- (x - DVLO)/(DVUP - DVLO + 1)
          xup <- (x - DVLO + 1)/(DVUP - DVLO + 1)
          cdflo <- pbeta(xlo, PMU*PHI, (1 - PMU)*PHI)
          cdfup <- pbeta(xup, PMU*PHI, (1 - PMU)*PHI)
          init + x*(cdfup - cdflo)
        })
      }
    )) %>%
  # Individual predictions and Pearson standard deviation
    mutate(tmp = pmap(list(MU, DVLO, DVUP, PHI), 
      .f = \(MU, DVLO, DVUP, PHI) {
        pmf <- map(seq(DVLO, DVUP), \(x) {
          xlo <- (x - DVLO)/(DVUP - DVLO + 1)
          xup <- (x - DVLO + 1)/(DVUP - DVLO + 1)
          cdflo = pbeta(xlo, MU*PHI, (1 - MU)*PHI)
          cdfup = pbeta(xup, MU*PHI, (1 - MU)*PHI)
          list(x = x, p = cdfup - cdflo)
        })
        tibble(IPRED = 0, PSD = 0) %>%
          reduce(.init = ., .x = pmf, .f = \(init, pmf) {
            mutate(init, IPRED = IPRED + pmf$p*pmf$x)
          }) %>%
          reduce(.init = ., .x = pmf, .f = \(init, pmf) {
            mutate(init, PSD = PSD + pmf$p*(pmf$x - IPRED)^2)
          }) %>%
          mutate(PSD = sqrt(PSD))
      }
    )) %>%
    unnest(tmp) %>%
  # Pearson weighted residuals
    mutate(PWRES = (DV - IPRED)/PSD)
```

```{r, echo = TRUE, warning = FALSE, message = FALSE, fig.dim = c(8, 8)}
# Create goodness of fit plots
  ipred_dv_plot <- obs_gof_plot(run9_fit, 
    pred_name = "IPRED", 
    dv_name = "DV", 
    cov_name = "PROJ",
    pred_label = "Individual Predicted")
  pred_dv_plot <- obs_gof_plot(run9_fit, 
    pred_name = "PRED", 
    dv_name = "DV",
    cov_name = "PROJ",
    pred_label = "Individual Predicted")
  wres_pred_plot <- res_gof_plot(run9_fit, 
    res_name = "PWRES", 
    x_name = "PRED",
    cov_name = "PROJ", 
    x_label = "Population Predicted",
    res_label = "Weighted Residual")
  wres_time_plot <- res_gof_plot(run9_fit, 
    res_name = "PWRES", 
    x_name = "TAFD",
    cov_name = "PROJ", 
    units = "hours", 
    x_label = "Time After First Dose",
    res_label = "Weighted Residual")
# Render plot
  combine_gof_plot(ipred_dv_plot, pred_dv_plot, 
    wres_pred_plot, wres_time_plot)
```

```{r, echo = TRUE, warning = FALSE, message = FALSE}
# Create individual plots
  run9_id_data <- mutate(run9_fit, SID = paste0("ID: ", ID))
  p <- individual_plots(run9_id_data, 
    dv_name = "DV",
    ipred_name = "IPRED",
    pred_name = "PRED",
    time_name = "TAFD",
    id_name = "SID",
    flag_name = "PROJ",
    plots_per_page = 6
  )
  grid.draw(p[[2]])
```

Again, this model could be entirely simulated in R, but an example using NONMEM 
is provided. Conducting simulations completely in NONMEM is more difficult for 
discretized beta regression. Therefore, NONMEM will be used to simulate the 
structural model (often desirable for more computationally challenging models 
containing differential equations), and R will be used to simulate from the 
discretized beta distribution. Simulation code is provided in 
`models/run9/sim9.mod`. Again, `.fit` and `.eta` files are pre-processed here 
to save on space, but also due to long runtime.

```{r, echo = TRUE, warning = FALSE, message = FALSE}
# Simulate using NONMEM structural model simulations
  # set.seed(555)
  # future::plan(future::multisession, workers = length(future::availableWorkers()) - 1)
  # run9_simeta <- file.path(run9_path, "sim9.eta") %>%
  #   read_table(na = c("NA", "***********", "1.#INFE+00"), skip = 1,
  #     col_types = cols(.default = col_double()))
  # run9_sim <- file.path(run9_path, "sim9.fit") %>%
  #   read_table(na = c("NA", "***********", "1.#INFE+00"), skip = 1,
  #     col_types = cols(.default = col_double())) %>%
  #   left_join(run9_eta, by = "ID") %>%  # Add subject level data
  #   filter(!is.na(SIM)) %>%
  #   mutate(PROJ = "PF2") %>%
  #   rename(MU = DV) %>%
  #   reduce2(.init = ., catcov_labels, cat_covs, \(init, cat, cov) {
  #     left_join(init, cat, by = cov)
  #   }) %>%
  #   mutate(DV = furrr::future_pmap_dbl(list(MU, PHI, DVLO, DVUP),
  #     .f = \(MU, PHI, DVLO, DVUP) {
  #       pmf <- accumulate(seq(DVLO, DVUP), .init = tibble(x = -99, p = 0), \(init, x) {
  #         xlo <- (x - DVLO)/(DVUP - DVLO + 1)
  #         xup <- (x - DVLO + 1)/(DVUP - DVLO + 1)
  #         cdflo = pbeta(xlo, MU*PHI, (1 - MU)*PHI)
  #         cdfup = pbeta(xup, MU*PHI, (1 - MU)*PHI)
  #         tibble(x = x, p = init$p + (cdfup - cdflo))
  #       })
  #       pmf %>%
  #         filter(runif(1) <= p) %>%
  #         head(1) %>%
  #         pull(x)
  #     }, .progress = TRUE, .options = furrr::furrr_options(seed = TRUE)
  #   ))
  # saveRDS(run9_sim, file.path(mod_dir, "run9", "sim9.rds"))
  run9_sim <- readRDS(file.path(mod_dir, "run9", "sim9.rds")) %>%
    mutate(PROJ = "PF2") %>%
    mutate(DV = 100*(1 - (DV/NQS - 1)/3))  # convert to score
  vpc_data <- list(obs = obs_data, sim = run9_sim) %>%
    map(mutate, WAFD = TAFD / 7)
  
# Define VPC bins
  tafd_labels <- c(0, 4, 8, 12, 16, 24, 40, 56)*7
  tafd_breaks <- c(-Inf, head(tafd_labels, -1) + diff(tafd_labels) / 2, Inf)
# Bin data for VPC
  tadbin_data <- vpc_data %>%
    map(bin_times, 
      time_var = "WAFD", 
      breaks = tafd_breaks, 
      labels = tafd_labels
    )
  
# Summarise and plot VPC for percent change from baseline
  vpc_summary <- tadbin_data %>%
    map(summarise_vpc,
      time_var = "TIMEBIN",
      facet_cov = "PROJ")
  plot_vpc(
    obs_summary = vpc_summary$obs,
    sim_summary = vpc_summary$sim,
    obs_data = tadbin_data$obs,
    time_var = "WAFD",
    time_label = "Time After First Dose (weeks)",
    dv_label = "Physical Functioning Score\n",
    png_name = "VPC_run9"
  )

```
